\specialchapt{ACKNOWLEDGEMENTS}

I would like to take a moment and thank everybody who has helped me get to this point. There are those that have helped me through the long night that is graduate school and those that helped convince me that graduate school was the right choice for me. I could write a thesis-length `thank you' section, but I will try to keep this brief.

To my advisors, Dr. Dan Nordman and Dr. Steve Vardeman, I want to thank you for being such wonderful mentors. I couldn't ask for more patient and understanding people to work with. Working with you has been an inspiration to me. I want to thank you for giving me such positive role models and showing me how to be a researcher, educator, and a statistician.

I would like to thank my committee members for their time and patience: Dr. Mark Kaiser, Dr. Kris DeBrabanter, and Dr. Dan Nettleton. I would also like to thank the Department of Statistics for encouraging a collaborative environment
and giving me a place to thrive. I have been very lucky to have mentors in the department in many capacities who should also be acknowledged. Thank you to Dr. Heike Hofmann, Dr. Ulrike Genschel, Dr. Dan Nettleton, and Dr. Di Cook for the guidance that you have given me in my tenure at Iowa State. I hope you know that I appreciated your time and mentorship greatly.

Finally, I would like to thank my friends and family for always believing in my success and providing me with the support to realize it. Thank you to my wife, Carrie Chennault. I simply don't know where I would be without you. You have helped make the last five years fly by with adventures and laughs. For my Dad, thank you for always lending an ear even when you have no idea what I'm talking about. I know you're proud of me, thank you for always telling me. To my friends, thank you for keeping me sane during this time. If we've shared a belay, bike ride, or beer, I'm talking to you.